Generative Simplex Mapping: Non-linear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery

### 2024-11-05 paper/paper.tex
Addedd patent section
updated for first round of review

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{We introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM). The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources. Points in this latent space are non-linearly mapped to reflectance spectra via a flexible function combining linear and non-linear mixing. Due to the probabilistic formulation of the GSM, spectral variability is also estimated by a precision parameter describing the distribution of observed spectra. Model parameters are determined using a generalized expectation-maximization algorithm. In the event of purely linear mixing, non-linear contributions are naturally driven to zero. The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic data set of linearly mixed spectra from the USGS spectral library. In a second experiment, the GSM is applied to real hyperspectral imagery captured over a pond in North Texas. The model is able to accurately identify spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source. Abundance maps generated using the GSM accurately track evolution of the dye plume as it mixes into the surrounding water.}

@@

The spatial resolution of hyperspectral imagers generally results in pixels with mixed signals from multiple sources called endmembers. The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction and abundance estimation. Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying HSI endmember spectra assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}. By further assuming a linear mixing model (LMM) in which observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed using a variety of techniques such as constrained least squares \cite{spectral-unmixing-orig, fcls-unmixing}. Among these methods, Nonnegative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}. The update equations for NMF can be formulated as multiplicative updates which guarantee the non-negativity of endmember spectra and their associated abundances \cite{nmf-algorithms}.  For this reason, the continued development of new NMF varieties remains an active area of research.

@@

The self-organizing map (SOM) is an unsupervised machine learning method that maps high-dimensional data to a low-dimensional grid while preserving the topological relationships between data points \cite{kohonen-som-1}. This low-dimensional representation provides a convenient way to visualize HSI data while the weight vectors for each SOM node can be interpreted as representative spectra \cite{cantero2004analysis, duran2007time,som-hsi}. If labelled reference spectra are available, the SOM can be used to enable semi-supervised labeling of HSI pixels \cite{riese2019supervised}. The SOM has also been shown to be effective for the compression of HSIs acquired by a CubeSat \cite{som-satellite}. Despite these capabilities, the SOM does not offer a probabilistic interpretation and relies on a heuristic training procedure with hyperparameters that can be challenging to tune. To address these shortcomings, Bishop et al. introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}. When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nodes to the HSI data space provides endmembers \cite{robot-team-gtm}. Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances, and the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.

In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures. By replacing the rectangular latent space of the GTM with a gridded $(n-1)$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n$ unique spectral signatures. The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex represent relative endmember abundances. Additionally, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra. If only linear mixing is present, the GSM algorithm drives non-linear contributions to $0$. Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of the resulting spectra and the degree of non-linear mixing applied.

@@

In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$. The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability. The GSM uses same structure, that is
\begin{equation}\label{eqn:data-space-distribution}
    p(\mathbf{x} \mid \mathbf{z}, \mathbf{W}, \beta) = \left(\frac{\beta}{2\pi} \right)^{D/2}\exp\left( -\frac{\beta}{2}\lVert \psi(\mathbf{z}; \mathbf{W}) - \mathbf{x} \rVert^2 \right)
\end{equation}
where $\mathbf{W}$ are model weights which parameterize the mapping $\psi$.

Assuming data are uniformly sampled from an embedded manifold in the data space, the GTM models the latent space using a rectangular grid with $K$-many nodes of equal prior probability. To adapt the GTM to describe endmember mixing, the GSM makes two key changes. The first is to replace the rectangular GTM grid with a gridded simplex with $N_v$ vertices. The barycentric coordinates for each node then describe the relative abundance of each endmember with each vertex corresponding to a pure endmember. The second change is to replace equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with nonuniform mixing distributions. Together, this leads to a latent space prior distribution given by
\begin{equation}\label{eqn:latent-distribution}
    p(\mathbf{z}) = \sum\limits_k^K \pi_k \, \delta(\mathbf{z} - \mathbf{z}_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1.
\end{equation}

@@

In the standard LMM model, the mapping $\psi$ is given by $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\mathbf{z}$ where the columns of $\mathbf{W}$ correspond to endmember spectra. To model non-linear mixing, $\mathbf{z}$ is replaced by the output of $M$-many activation functions such that $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\phi(\mathbf{z})$. The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(\mathbf{z}_k)$. For $m \leq N_v$, we take $\Phi_{km} = [\mathbf{z}_k]_m$ to model linear mixing. The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) with 
\begin{equation}\label{eq:act-function}
    \Phi_{km} = \begin{cases}
        \dfrac{s - \lVert \mathbf{z}_k - \mu_m \rVert}{s}; & \lVert \mathbf{z}_k - \mu_m \rVert \leq s \\
        0; & \lVert \mathbf{z}_k - \mu_m \rVert > s
    \end{cases}
\end{equation}
where $s$ is the spacing between RBF centers. In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra, while the remaining columns account for additional non-linear effects. For linear mixing, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$. A visualization of the GSM model is shown in Figure~\ref{fig:gsm-diagram}.

\begin{figure}[H]
\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.png}
\caption{Illustration of the GSM. The latent space consists of a grid of $K$-many points (green dots) distributed throughout a simplex with $N_v$ vertices. Barycentric coordinates of each node in the simplex correspond to the relative abundance of $N_v$-many unique sources. Here, $N_v=3$ has been chosen for illustrative purposes. Nodes are mapped into the data space via the map $\psi(\mathbf{z})$ utilizing $M$-many radially symmetric basis functions (red). Spectral variability is estimated via the precision parameter $\beta$ shown here in the data space as a light blue band around the spectrum given by $\psi(\mathbf{z})$.}
\label{fig:gsm-diagram}
\end{figure}

@@

An EM algorithm for the GSM model can now be formulated as follows. Suppose that we have current estimates for the model weights $\mathbf{W}$, mixing coefficients $\pi_k$, and precision parameter $\beta$. During the expectation step we compute the posterior probabilities, that is, the responsibility of each GSM node for each spectrum in the data set:
\begin{equation}\label{eq:responsibility}
    R_{kn}  = p(\mathbf{z}_k \mid \mathbf{x}_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(\mathbf{x}_n \mid \mathbf{z}_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(\mathbf{x}_n \mid \mathbf{z}_{k'}, \mathbf{W}, \beta)}.
\end{equation}
For the maximization step, we consider the expectation of the penalized complete-data log likelihood given by
\begin{equation}\label{eq:complete-data-llh}
\begin{aligned}
    Q &= \sum_n^N\sum_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum_d^D\left(\sum_m^M W_{dm}\Phi_{km} - X_{nd}\right)^2\right) \\ 
    &\qquad + \frac{N_vD}{2}\ln\left(\frac{\lambda_e}{2\pi}\right) - \frac{\lambda}{2}\sum_d^D \sum_{m=1}^{N_v} W_{dm}^2  \\ 
    &\qquad + (M-N_v)D\ln\left(\dfrac{\lambda_w}{2}\right) - \lambda_w\sum_d^D\sum_{m=N_v+1}^{M} W_{dm}
\end{aligned}
\end{equation}
where $X_{nd}$ is the $d$-th component of the $n$-th spectrum in the data set. Eq~\ref{eq:complete-data-llh} is then maximized with respect to $\pi_k$, $\beta$, and $\mathbf{W}$ to obtain new parameter values. For a detailed overview of the EM procedure, the reader is directed to ref. \cite{bishop-prml}.

@@

To illustrate the effectiveness of the GSM, we first demonstrate its ability to model simple linear mixing by driving non-linear weights to zero during the fitting process. To this end, a synthetic data set comprising linear mixtures of three sample spectra from the U.S. Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}. These source spectra and abundances are visualized in Figure~\ref{fig:usgs-data}.

\begin{figure}[H]
\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
\caption{Synthetic data set formed from USGS spectra. \textbf{(a)} Spectra from the USGS spectral database used as the ground truth endmembers. These spectra were selected following the example in ref. \cite{vca-orig}. \textbf{(b)} The abundance distribution sampled for in the data set. Samples were generated from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
\label{fig:usgs-data}
\end{figure}

@@

For comparison, NMF models were also trained on each data set as this method is both highly popular and does not include the pure-pixel assumption common to other technqieus like VCA and PPI. Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF variations \cite{unmixing-nmf-review}. For the purpose of evaluating the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}. We also included the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}.

@@

The quality of the unmixing, that is, the estimation of the abundance, performed by each model was evaluated using the mean abundance RMSE. Here we again see that both versions of the GSM outperformed NMF. To justify that all models were fairly trained to convergence, the RMSE data reconstruction was also computed. This metric uses the trained mixing model to compute the error between the original data set and the reconstructed spectra generated using the extracted endmembers and their abundances. For all GSM and NMF models, the RMSE data reconstruction converged to the level of random noise introduced into the data. These results reflect a fair comparison between the NMF and GSM models.

@@

%\section{Patents}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{We introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM). The model represents endmember mixing using a latent space of points sampled within a $(n-1)$-simplex corresponding to $n$ unique sources. Barycentric coordinates within this simplex are naturally interpreted as relative endemember abundances satisfying both the abundance sum-to-one and abundance non-negativity constraints. Points in this latent space are mapped to reflectance spectra via a flexible function combining linear and non-linear mixing. Due to the probabilistic formulation of the GSM, spectral variability is also estimated by a precision parameter describing the distribution of observed spectra. Model parameters are determined using a generalized expectation-maximization algorithm which guarantees non-negativity for extracted endmembers. We first compare the GSM against three varieties of nonnegative matrix factorization (NMF) on a sythetic dataset of linearly mixed spectra from the USGS spectral database. Here the GSM performed favorably for both endmember accacy and abundance estimation with the all non-linear contributions driven to zero by the fitting procedure. In a second experiment, we apply the GTM to model non-linear mixing in real hyperspectral imagery captured over a pond in North Texas. The model accurately identified spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source. Abundance maps generated using the GSM accurately track the evolution of the dye plume as it mixes into the surrounding water.}

% In the event of purely linear mixing, non-linear contributions are naturally driven to zero. The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic data set of linearly mixed spectra from the USGS spectral library. In a second experiment, the GSM is applied to real hyperspectral imagery captured over a pond in North Texas. The model is able to accurately identify spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source. Abundance maps generated using the GSM accurately track evolution of the dye plume as it mixes into the surrounding water.}

@@

The spatial resolution of hyperspectral imagers generally results in pixels with mixed signals from multiple sources called endmembers. The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction and abundance estimation. Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying HSI endmember spectra assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}. By further assuming a linear mixing model (LMM) in which observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed using a variety of techniques such as constrained least squares and multi-objective optimization \cite{spectral-unmixing-orig, fcls-unmixing, gmogh}. Among these methods, Nonnegative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}. The update equations for NMF can be formulated as multiplicative updates which guarantee the non-negativity of endmember spectra and their associated abundances \cite{nmf-algorithms}.  For this reason, the continued development of new NMF varieties remains an active area of research.

@@

The self-organizing map (SOM) is an unsupervised machine learning method that maps high-dimensional data to a low-dimensional grid while preserving the topological relationships between data points \cite{kohonen-som-1}. This low-dimensional representation provides a convenient way to visualize HSI data while the weight vectors for each SOM node can be interpreted as representative spectra \cite{cantero2004analysis, duran2007time,som-hsi}. If labelled reference spectra are available, the SOM can be used to enable semi-supervised labeling of HSI pixels \cite{riese2019supervised}. The SOM has also been shown to be effective for the compression of HSIs acquired by a CubeSat \cite{som-satellite}. Despite these capabilities, the SOM does not offer a probabilistic interpretation and relies on a heuristic training procedure with hyperparameters that can be challenging to tune. To address these shortcomings, Bishop et al. introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}. When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nods to the HSI data space provides endmembers \cite{robot-team-gtm}. Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances, and the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.

In this paper, we introduce a new variant of the GTM, dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures. By replacing the rectangular latent space of the GTM with a gridded $(n-1)$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n$ unique spectral signatures. The mapping from the latent space to the HSI data space accounts for linear and non-linear spectral mixing while barycentric coordinates for the latent space simplex represent relative endmember abundances. Additionally, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra. If only linear mixing is present, the GSM algorithm drives non-linear contributions to $0$. Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of the resulting spectra and the degree of non-linear mixing applied. In summary, the key innovations introduced by the GSM are:
\begin{itemize}
    \item the GSM can model linear \textit{and} nonlinear spectral mixing
    \item the GSM does not assume the presence of pure pixels in the dataset 
    \item the probabilistic formulation of the GSM accounts for spectral variability
    \item the simplex used for the latent space structure of the GSM is directly interpretable and forces abundances to satisfy both the abundance sum-to-one and abundance non-negativity constraints
    \item the fiting procedure introduced for the GSM maintains non-negativity of endmember spectra
\end{itemize}

%  \textcolor{red}{ In summary, the key innovations of the GSM are
% \begin{itemize}
%     \item Modeling of linear and non-linear spectral mixing
%     \item The model is probabilistic, with spectral varaibility accounted for via a precision parameter
%     \item Satisfaction of constraints
%     \item No pure-pixel assumption
% \end{itemize}    
% }

@@

In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra of length $D$) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$. The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability. The GSM uses same structure, that is
\begin{equation}\label{eqn:data-space-distribution}
    p(\mathbf{x} \mid \mathbf{z}, \mathbf{W}, \beta) = \left(\frac{\beta}{2\pi} \right)^{D/2}\exp\left( -\frac{\beta}{2}\lVert \psi(\mathbf{z}; \mathbf{W}) - \mathbf{x} \rVert^2 \right)
\end{equation}
where $\mathbf{z}=(z_1, z_2, ..., z_{N_v})$ corresponding to $N_v$-many sources and $\mathbf{W}$ is a $D \times M$ matrix of model weights which parameterize the mapping $\psi$.

Assuming data are uniformly sampled from an embedded manifold in the data space, the GTM models the latent space using a rectangular grid with $K$-many nodes of equal prior probability. To adapt the GTM to describe endmember mixing, the GSM makes two key changes. The first is to replace the rectangular GTM grid with a gridded simplex with $N_v$ vertices. Barycentric coordinates for each node then describe the relative abundance of each endmember with siplex vertices corresponding to pure sources. The second change is to augment the equal prior probabilities of the GTM latent space with adaptive mixing coefficients $\pi_k$ to allow for nonuniform sampling across the abundance simplex. This addition allows the GSM to flexibly address many possible mixing scenarios without an \textit{a priori} assumption for a specific mixing distribution. Together, these changes result in a prior distribution for the latent space given by
\begin{equation}\label{eqn:latent-distribution}
    p(\mathbf{z}) = \sum\limits_k^K \pi_k \, \delta(\mathbf{z} - \mathbf{z}_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1.
\end{equation}
where $\delta(\cdot)$ is the Dirac delta function and $\mathbf{z}_k$ are the positions of each node within the simplex.

@@

In the standard LMM model, the mapping $\psi$ is given by $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\mathbf{z}$ where the columns of $\mathbf{W}$ correspond to endmember spectra. To model non-linear mixing, $\mathbf{z}$ is replaced by the output of $M$-many activation functions such that $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\phi(\mathbf{z})$. The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(\mathbf{z}_k)$. For $m \leq N_v$, we take $\Phi_{km} = [\mathbf{z}_k]_m$ to model linear mixing. The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) with 
\begin{equation}\label{eq:act-function}
    \Phi_{km} = \begin{cases}
        \dfrac{s - \lVert \mathbf{z}_k - \mu_m \rVert}{s}; & \lVert \mathbf{z}_k - \mu_m \rVert \leq s \\
        0; & \lVert \mathbf{z}_k - \mu_m \rVert > s
    \end{cases}
\end{equation}
where $s$ is the spacing between RBF centers. A visual representation of the non-linear activation functions in Equation~\ref{eq:act-function} is shown below in Figure~\ref{fig:act-func} for a 2-component GSM.
\begin{figure}[!b]
    \centering
    \includegraphics[width=0.45\columnwidth]{methods/gsm/activation-function.pdf}
    \caption{Illustration of the non-linear activation functions for a 2-component GSM. The simplex vertices are shown as green dots ($\mu_1$ and $\mu_2$) and three RBF centers are shown in red ($\mu_1$, $\mu_2$, and $\mu_3$). With no RBF centers at the vertices, we guarantee no non-linear contributions occur for pure endmember spectra.}
    \label{fig:act-func}
\end{figure}
In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra, while the remaining columns account for additional non-linear effects. Equation~\ref{eq:act-function} is specifically chosen so that no non-linear contributions are possible for pure spectra at the vertices of the simplex. At all other points, the output of $\psi$ involves both linear and non-linear contributions. If only linear mixing is present, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$. A visualization of the GSM model for three sources is shown in Figure~\ref{fig:gsm-diagram}.

\begin{figure}
    \centering
\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.png}
\caption{Illustration of the GSM. The latent space consists of a grid of $K$-many points (green dots) distributed throughout a simplex with $N_v$ vertices. Barycentric coordinates of each node in the simplex correspond to the relative abundance of $N_v$-many unique sources. Here, $N_v=3$ has been chosen for illustrative purposes. Nodes are mapped into the data space via the map $\psi(\mathbf{z})$ utilizing $M$-many radially symmetric basis functions (red). Spectral variability is estimated via the precision parameter $\beta$ shown here in the data space as a light blue band around the spectrum given by $\psi(\mathbf{z})$.}
\label{fig:gsm-diagram}
\end{figure}

@@

An EM algorithm for the GSM model can now be formulated as follows. Suppose that we have current estimates for the model weights $\mathbf{W}$, mixing coefficients $\pi_k$, and precision parameter $\beta$. During the expectation step we compute the posterior probabilities, that is, the responsibility of each GSM node for each spectrum in the data set:
\begin{equation}\label{eq:responsibility}
    R_{kn}  = p(\mathbf{z}_k \mid \mathbf{x}_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(\mathbf{x}_n \mid \mathbf{z}_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(\mathbf{x}_n \mid \mathbf{z}_{k'}, \mathbf{W}, \beta)}.
\end{equation}
For the maximization step, we consider the expectation of the penalized complete-data log likelihood given by
\begin{equation}\label{eq:complete-data-llh}
\begin{aligned}
    Q &= \sum_n^N\sum_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum_d^D\left(\sum_m^M W_{dm}\Phi_{km} - X_{nd}\right)^2\right) \\ 
    &\qquad + \frac{N_vD}{2}\ln\left(\frac{\lambda_e}{2\pi}\right) - \frac{\lambda_e}{2}\sum_d^D \sum_{m=1}^{N_v} W_{dm}^2  \\ 
    &\qquad + (M-N_v)D\ln\left(\dfrac{\lambda_w}{2}\right) - \lambda_w\sum_d^D\sum_{m=N_v+1}^{M} W_{dm}
\end{aligned}
\end{equation}
where $X_{nd}$ is the $d$-th component of the $n$-th spectrum in the data set. Eq~\ref{eq:complete-data-llh} is then maximized with respect to $\pi_k$, $\beta$, and $\mathbf{W}$ to obtain new parameter values. For a detailed overview of the EM procedure, the reader is directed to ref. \cite{bishop-prml}.

@@

To illustrate the effectiveness of the GSM, we first demonstrate its ability to model linear mixing. This serves as an important limiting case since linearly mixed spectra should not lead to the spurous introduction of non-linear contributions by the GSM. The goal of this first test is therefore to demonstrate that the GSM drives non-linear weights to zero for linearly mixed data while providing a fair test to compare GSM to well-established linear mixing models. To this end, a synthetic data set comprising linear mixtures of three sample spectra from the U.S. Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}. These source spectra and abundances are visualized in Figure~\ref{fig:usgs-data}.

\begin{figure}[H]
\includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
\caption{Synthetic data set formed from USGS spectra. \textbf{(a)} Spectra from the USGS spectral database used as the ground truth endmembers. These spectra were selected following the example in ref. \cite{vca-orig}. \textbf{(b)} The abundance distribution sampled for in the data set. Samples were generated from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
\label{fig:usgs-data}
\end{figure}

@@

For comparison, three NMF models were also trained on each data set as this method is both highly popular and does not include the pure-pixel assumption common to other techniques like VCA and PPI. Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF varieties \cite{unmixing-nmf-review}. For the purpose of comparing to the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}. We also included the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}. We note that the goal of this test is not to prove the GSM is superior to other models for linear mixing, but rather to demonstrate that the GSM can model linearly mixed data without introducing unnecessary complexity.

@@

The quality of the unmixing, that is, the estimation of the abundance, performed by each model was evaluated using the mean abundance RMSE. Here we again see that both versions of the GSM outperformed NMF. To justify that all models were fairly trained to convergence, the data reconstruction RMSE was also computed. This metric uses the trained mixing model to compute the error between the original data set and the reconstructed spectra generated using the extracted endmembers and their abundances. For all GSM and NMF models, the data reconstruction RMSE converged to the level of random noise introduced into the data. These results reflect a fair comparison between the NMF and GSM models.

@@

\section{Patents}
A provisional US patent application for this work has been filed on 2024-10-10 entitled: \textit{Generative Simplex Mapping: Non-Linear Endmember Extraction And Spectral Unmixing For Hyperspectral Imagery}, application number 63/705,854.
